Verifying Hierarchical Cache Coherence Protocols: a theorem-prover based modular, reusable proof

### 2014-02-05 paper/Intuition.tex
A pass over Intuition

\section{Proof of MSI protocol being preserving Store atomicity}
\label{Sec:Safety}

In this section, we present the structure of the proof of Theorem
\fullref{storeAtomicity} for a system implementing the MSI protocol of
Figure~\ref{trans}. We reduce the whole system proof to that of a set
of invariants between any parent-child pair. These invariants in turn
depend on two smaller invariants which encapsulate the core aspects of
correctness of the MSI protocol and allow the large scope of verified
hardware refinements we present later (Section~\ref{Sec:Refinements}).

The invariants hold after any number of executions of the transactions
starting from the initial state. We use the term time $t$ to specify
the global state reached after firing $t$ transitions.

While we have proofs in Coq for each invariant our discussion will be mainly
informal for brevity. A full description can be found at~\xxx{MURALI PUT
REFERENCE HERE}.

\subsection{Reducing correctness to Node-level Invariants}

Store atomicity is a whole system property. Intuitively our
protocol is correct because it guarantees at most one cache may
write at a time, and that any cache permitted to read has
the ``latest'' value. We state this formally with the following 4
invariants.

\begin{inv}
\textit{latestValue}:
At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
up-to-date value for that address, \ie{}
\begin{enumerate}
\item $c.\data[a] = \initData(a)$ and no store request $q'$ has been processed at
any time $t'$ such that $t' < t \wedge a = q'.\addrQ$, or
\item There is a store request $q$ that was processed at time $t_q$ such that
$t_q < t \wedge a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
request $q'$ was processed at any time $t'$ such that $t_q < t' < t \wedge a =
q'.\addrQ$.
\end{enumerate}
\label{latestValue}
\end{inv}
A clean cache is defined as follows.
\begin{defn}
\textit{clean}: A cache is said to be \clean{} for an address $a$ if and only if
the state of the cache for that address is at least \Sh{} and the directory 
state for all its children is at most \Sh.
\end{defn}

\begin{inv}
\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
that neither is an ancestor of the other in the cache tree
hierarchy (where ancestor relation is the reflexive-transitive closure of the parent relation), for each
address $a$, $c_1.\state[a]$ and $c_2.\state[a]$ are \textit{compatible}.
\label{nonAncestorCompatible}
\end{inv}
Compatibility of states is defined as follows.
\begin{defn}
\textit{compatibile}: Values of states $x_1$ and $x_2$ are \textit{compatible} iff
either $x_1$ or $x_2$ is \Mo{}, then the other is \In. Note that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also compatible if $x_1, x_2$ are compatible.
\label{compatible}
\end{defn}

@@

Given these invariants, we codify our intuition by showing that
Theorem~\fullref{storeAtomicity} follows naturally from these
invariants.  A store may happen only when leaf cache is in \Mo{} state
for that address (Invariant~\fullref{processReq}). Leaf caches cannot
be ancestors of any other node, so by Invariant
\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all
other leaf caches will be in \In{} state preventing the possibility of
load and store requests for the same address (by Invariant
\fullref{processReq}), showing Definition~\fullref{uniqStTimes} holds.

Given Invariant~\fullref{noReqAgain}, once a response has been produced
for a request $i$, the cache may subsequently only process $j^{th}$
request such that $j > i$. Thus there is at most one response to each
request, satisfying Definition~\fullref{uniqRespLabels}. A load request
may be processed only if a cache is at least in \Sh{} state (Invariant
\fullref{processReq}). As leaf caches have no children, a leaf cache in
at least in \Sh{} state for address $a$ is \clean for that address. A
load response supplies the data for that address present in the cache,
which, by Invariant~\fullref{lastestValue}, contains the ``latest
value''. Thus, the final condition for Theorem~\fullref{storeAtomicity}
is also satisfied. All that remains is to show that these properties hold.

\subsection{Reducing to Pair-wise Invariants}
\label{pairwise}

@@

For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
the cache tree hierarchy, from Invariants \fullref{conservative} and
\fullref{localCompatibility}, we can show that for any address $a$,
$c.\state[a] \le p.\state[c][a]$. This is because the ancestor
relation is the reflective-transitive closure of the parent relation,
and Invariants \fullref{conservative} and the first part
\fullref{localCompatibility} of defines a $\ge$ relation between the
states of a parent and its child.

@@

We prove Invariant \fullref{latestValue} by strong induction on
time. Consider two cases: cache $c$ is already \clean{} for an address
$a$ at time $t$ or the contrary.  If \clean{}, then it has the latest
value at time $t$. Since none of its children's directory entries can
be in \Mo{}, all its descendants should be no greater than \Sh{} for
that address (as discussed above). Any non-ancestor of $c$, and
non-descendant of $c$ must be compatible (as discussed above). Since
$c$ is \clean, it is at least in the \Sh{} state, so compatibility
means that non-descendants can not be in \Mo{} state. All of $c$'s
ancestors are non-leaves, so they can not process any store
request. Therefore, no cache save $c$ may have process a store
request, which means that $c$ contains the latest value for address
$a$ at time $t+1$ if it was \clean{} at time $t$.

If $c$ was not clean at time $t$, it must have received data from
either its parent or its child.  As can be seen in Figure~\ref{trans},
whenever data is sent either from a parent to a child, or from a child
to the parent, the sender is \clean. By the induction hypothesis, it
means that the sender has the latest value for the address. Because of
Invariant~\ref{noTransitWrite}, no store request can be processed when
the data is in transit, which means $c$ will have the latest value for
address $a$ at time $t+1$.

\subsection{Proving parent-child hierarchy properties}

@@

@@

\begin{inv}
\textit{noCross}: A response from a cache to its child for an address can not
be in transit while another response from the same child to the parent for the
same address is in transit.
\label{noCross}
\end{inv}

@@

To prove Invariant \fullref{conservative} at some time $t$, consider the case
when a child cache $c$ has sent to or received from the parent $p$ some message
before $t$. Consider the last such message $m$ sent by or received at $c$ at
$t'$. Let's say it is sent by $c$ at $t'$, not received. Then $p$ can not have
received any message from $c$ after sending $m$ and before $t$ in order to not
violate Invariant \fullref{respFIFO}.  Instead, if $c$ receives $m$ at $t'$,
$p$ will not receive any message after sending $m$ and before $t$ in order to
not violate Invariant \fullref{noCross}. Thus, $p$ will not receive any message
from $c$ after sending or receiving $m$ and before $t$.  Let $t''$ be the time
at which $p$ has sent or received $m$. Whether $p$ sends or receives $m$,
$p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ because
the response messages sync up the next-state according to the transitions in
Figure \ref{trans}. Since $p$ can not receive any message between $t''+1$ and
$t$, it can only upgrade $p.\dstate[c][a]$. $c.\state[a]$ does not change
between $t'+1$ and $t$ as it never changes state. Therefore, Invariant
\fullref{conservative} holds at $t$.

If no message has been sent or received by $c$ before $t$, then $p$ wouldn't
have received any message sent by $c$ before $t$. So $p.\dstate[c][a]$ at time
$t \ge p.\dstate[c][a]$ at time $0$ . Since the initial values of the directory
and the child's state are both set to \In{} Invariant \textit{conservative}
holds.

@@

If the value of the state of a child at the time of sending a response is the
same as the value of the directory for that child at the parent at the time of
receiving the response, it follows straightaway that the parent downgrades its
directory on receiving a response from the child, since a child downgrades on
sending a response and the next-states are synced on receiving the response.
The proof for this follows exactly the same arguments presented above to prove
Invariant \fullref{conservative} -- we prove by considering the last message
sent or received at the child before $t$ (and the case when no message has been
sent or received).

>>>>

\section{Proof of Store Atomicity for the MSI Protocol}
\label{Sec:Safety}

In this section, we present the structure of the proof of Theorem
\fullref{storeAtomicity} for a system implementing the MSI protocol of
Figure~\ref{trans}. We reduce the whole system proof to that of a set
of invariants between any parent-child pair. These invariants in turn
depend on two smaller invariants that encapsulate the core aspects of
correctness of the MSI protocol and allow the large scope of verified
hardware refinements we present later (Section~\ref{Sec:Refinements}).

The invariants hold after any number of executions of the transactions
starting from the initial state. We write ``time $t$'' for
the global state reached after firing $t$ transitions.

While we have proofs in Coq for each invariant, our discussion will be mainly
informal for brevity. A full description can be found at~\xxx{MURALI PUT
REFERENCE HERE}.
\adamc{I presume we'll want to submit a tarball of the Coq source code with the paper.
Is that what you have in mind here?}

\subsection{Reducing Correctness to Node-Level Invariants}

Store atomicity is a whole-system property. Intuitively our
protocol is correct because it guarantees at most one cache may
write at a time, and that any cache permitted to read has
the ``latest'' value. We formalize this intuition with the following 4
invariants.

\begin{inv}
\textit{latestValue}:
At any time $t$, a cache $c$ that is clean for an address $a$ will have the most
up-to-date value for that address, \ie{}
\begin{enumerate}
\item $c.\data[a] = \initData(a)$ and no store request $q'$ has been processed at
any time $t'$ such that $t' < t \wedge a = q'.\addrQ$, or
\item There is a store request $q$ that was processed at time $t_q$ such that
$t_q < t \wedge a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
request $q'$ was processed at any time $t'$ such that $t_q < t' < t \wedge a =
q'.\addrQ$.
\end{enumerate}
\label{latestValue}
\end{inv}
A clean cache is defined as follows.
\begin{defn}
\textit{clean}: A cache is said to be \clean{} for an address $a$ if and only if
the state of the cache for that address is at least \Sh{} and the directory 
state for all its children is at most \Sh.
\end{defn}

\begin{inv}
\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
that neither is an ancestor of the other in the cache tree
hierarchy (where the ancestor relation is the reflexive-transitive closure of the parent relation), for each
address $a$, $c_1.\state[a]$ and $c_2.\state[a]$ are \textit{compatible}.
\label{nonAncestorCompatible}
\end{inv}
Compatibility of states is defined as follows.
\begin{defn}
\textit{compatibile}: Values of states $x_1$ and $x_2$ are \textit{compatible} iff
whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In. Note that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also compatible if $x_1, x_2$ are compatible.
\label{compatible}
\end{defn}

@@

\adamc{Something needs to be fixed up here, for consistency in how many invariants there are.}

Given these invariants, we codify our intuition by showing that
Theorem~\fullref{storeAtomicity} follows naturally from the
invariants.  A store may happen only when a leaf cache is in the \Mo{} state
for that address (Invariant~\fullref{processReq}). Leaf caches cannot
be ancestors of any other node, so by Invariant
\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all
other leaf caches will be in \In{} state preventing the possibility of
load and store requests for the same address (by Invariant
\fullref{processReq}), showing Definition~\fullref{uniqStTimes} holds.

Given Invariant~\fullref{noReqAgain}, once a response has been produced
for a request $i$, the cache may subsequently only process a $j^{th}$
request such that $j > i$. Thus there is at most one response to each
request, satisfying Definition~\fullref{uniqRespLabels}. A load request
may be processed only if a cache is at least in \Sh{} state (Invariant
\fullref{processReq}). As leaf caches have no children, a leaf cache in
at least the \Sh{} state for address $a$ is \clean{} for that address. A
load response supplies the data for that address present in the cache,
which, by Invariant~\fullref{lastestValue}, contains the ``latest
value.'' Thus, the final condition for Theorem~\fullref{storeAtomicity}
is also satisfied. All that remains is to show that the system maintains the
invariants.

\subsection{Reducing to Pairwise Invariants}
\label{pairwise}

@@

For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
the cache tree hierarchy, from Invariants \fullref{conservative} and
\fullref{localCompatibility}, we can show that for any address $a$,
$c.\state[a] \le p.\state[c][a]$. The reason is that the ancestor
relation is the reflective-transitive closure of the parent relation,
and Invariants \fullref{conservative} and the first part
of \fullref{localCompatibility} imply a $\ge$ relation between the
states of a parent and its child.

@@

We prove Invariant \fullref{latestValue} by strong induction on
time. Consider two cases: cache $c$ is already \clean{} for an address
$a$ at time $t$ or the contrary.  If \clean{}, then it has the latest
value at time $t$. Since none of its children's directory entries can
be in \Mo{}, all its descendants should be no greater than \Sh{} for
that address (as discussed above). Any non-ancestor of $c$ and
non-descendant of $c$ must be compatible (as discussed above). Since
$c$ is \clean, it is at least in the \Sh{} state, so compatibility
means that non-descendants cannot be in \Mo{} state. All of $c$'s
ancestors are non-leaves, so they cannot process any store
request. Therefore, no cache save $c$ may have processed a store
request, which means that $c$ contains the latest value for address
$a$ at time $t+1$ if it was \clean{} at time $t$.

If $c$ was not clean at time $t$, it must have received data from
either its parent or its child.  As can be seen in Figure~\ref{trans},
whenever data is sent either from a parent to a child or from a child
to the parent, the sender is \clean. By the induction hypothesis, it
means that the sender has the latest value for the address. Because of
Invariant~\ref{noTransitWrite}, no store request can be processed when
the data is in transit, which means $c$ will have the latest value for
address $a$ at time $t+1$.

\subsection{Proving Parent-Child Hierarchy Properties}

@@

\adamc{This section seems like a prime place to reduce detail to save space, while also saving readers from trying to understand proofs that are mechanized, anyway.}

@@

\begin{inv}
\textit{noCross}: A response from a cache to its child for an address cannot
be in transit while another response from the same child to the parent for the
same address is in transit.
\label{noCross}
\end{inv}

@@

To prove Invariant \fullref{conservative} at some time $t$, consider the case
when a child cache $c$ has sent to or received from the parent $p$ some message
before $t$. Consider the last such message $m$ sent by or received at $c$ at
$t'$. If it is sent rather than received by $c$ at $t'$, then $p$ cannot have
received any message from $c$ after sending $m$ and before $t$ in order to not
violate Invariant \fullref{respFIFO}.  Instead, if $c$ receives $m$ at $t'$,
$p$ will not receive any message after sending $m$ and before $t$ in order to
not violate Invariant \fullref{noCross}. Thus, $p$ will not receive any message
from $c$ after sending or receiving $m$ and before $t$.  Let $t''$ be the time
at which $p$ has sent or received $m$. Whether $p$ sends or receives $m$,
$p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ because
the response messages sync up the next-state according to the transitions in
Figure \ref{trans}. Since $p$ cannot receive any message between $t''+1$ and
$t$, it can only upgrade $p.\dstate[c][a]$. $c.\state[a]$ does not change
between $t'+1$ and $t$ as it never changes state. Therefore, Invariant
\fullref{conservative} holds at $t$.

If no message has been sent or received by $c$ before $t$, then $p$ would not
have received any message sent by $c$ before $t$. So $p.\dstate[c][a]$ at time
$t$ is greater than $p.\dstate[c][a]$ at time $0$. Since the initial values of the directory
and the child's state are both set to \In{}, Invariant \textit{conservative}
holds.

@@

If the value of the state of a child at the time of sending a response is the
same as the value of the directory for that child at the parent at the time of
receiving the response, it follows straightaway that the parent downgrades its
directory on receiving a response from the child, since a child downgrades on
sending a response and the next-states are synced on receiving the response.
The proof for this follows exactly the same arguments presented above to prove
Invariant \fullref{conservative} -- we prove it by considering the last message
sent or received at the child before $t$ (and the case when no message has been
sent or received).
